---
title: "piCurve Package"
output:
  md_document:
    variant: markdown_github
---

## Overview

The piCurve package offers a comprehensive suite of photosynthesis–irradiance (PI) models in a user-friendly environment, enabling users to explore how varying irradiance levels influence photosynthetic or growth rates using statistical methods. The models have been rigorously tested and validated against experimental data, ensuring both reliability and accuracy. In total, 24 PI models are included this library (`?Model_piCurve`). Optimal parameters for each model formulation are estimated through non-linear optimization using two statistical approaches: mean squared error (MSE) and maximum likelihood estimation (MLE). A built-in dataset (`?piDataSet`) containing eight independent PI incubation samples is included for model testing and validation.

In addition to PI-curve fitting, **piCurve package now provides tools for radiative transfer and optical modelling**, including:

- Estimation of **downwelling spectral irradiance** at the sea surface using a Gregg–Carder–type parameterization with climatological atmospheric inputs.
- Computation of **photosynthetically available radiation (PAR)** by spectrally integrating downwelling irradiance.
- Estimation of **phytoplankton absorption coefficients** from chlorophyll-a concentration using established bio-optical parameterizations.

These capabilities enable users to link physiological photosynthesis models with environmental forcing, facilitating analyses ranging from single-experiment PI curve fitting to regional or global assessments of marine primary productivity.

## Installation
 
Install the package from GitHub:
 
```{r, message=FALSE}
remotes::install_github("Mohammad-Amirian/piCurve")
```


## How to use piCurve

```{r load, message=FALSE}
library(piCurve)
library(ggplot2)
library(dplyr)
library(maps)
library(grid)
```


```{r, echo=FALSE, message=FALSE}
# library(viridis)
# library(fBasics)
```

### Classfying Data

To classify PI data as light-limited (ll), light-saturated (ls), or photoinhibited (ph), you can use the `DataType_piCurve()` function. Below, we demonstrate how to apply this function to the built-in dataset.

```{r}
# Split the data
grouped <- 
    piDataSet |> 
    dplyr::group_by(pi_number) |> 
    dplyr::group_split()

# Extract pi_number values
pi_numbers <- 
    piDataSet |> 
    dplyr::group_by(pi_number) |> 
    dplyr::group_keys() |> 
    dplyr::pull(pi_number)

# Apply function and combine with pi_number
result_df <- data.frame(
    pi_number = pi_numbers,
    data_type = unlist(parallel::mclapply(grouped, function(x) DataType_piCurve(data = x, n_cores = 1)$data_type, mc.cores = 6))
)
 
dplyr::count(result_df, data_type)
```

# Model Fitting

To fit a PI model to a given data sample, you can use the `Fit_piModel()` function. In the example below, we use the default settings of this function, which fits the Amirian–tanh model (Ph10; see `?Model_piCurve()`) to the data, assuming a dark reaction rate of zero (R = 0). This model is chosen because it provides coherent and consistent outputs. For more details, please refer to the associated reference.

```{r}
df <- grouped[[5]]

# fit model on a given data
fit <- Fit_piModel(data = df, STATapp = "MLE", Hessian = TRUE)

print(fit)
```

### Tidy Fit

You can use the `Tidy_piCurve()` function to convert the output of your model fit into a tidy tibble format.

```{r}
# tidy the output of the model
Tidy_piCurve(fit)$params
```

### Plotting

You can use the `Plot_piCurve()` function to visualize the fitted model output along with the predicted interval (95% by default) over the data.

```{r}
p <- # plot the fitted model 
    Plot_piCurve(fit, df, length_out = 200, add_CI = TRUE, n_cores = 6)

p + # increase the size of axis title and font
    theme(axis.text = element_text(size = 14), axis.title = element_text(size = 14))
```

### Other Useful Functions

In addition to the core functions described above, the piCurve package includes several other helpful tools for model analysis and interpretation. A few of them are highlighted below.

If the user forgot to set `Hessian = TRUE` when using the `Fit_piModel()` function, the information matrix can still be calculated afterward using the `InfoMat_piCurve()` function, as shown below:

```{r}
InfoMat_piCurve(parameters = fit$par, model_name = fit$model, data = df)
```

To add confidence intervals to the predicted response variable (`y`), you can use the `addCI_to_piPred()` function.

```{r}
addCI_to_piPred(Fitted_Model = fit, irrad = df$I) |> head(3)
```

To increase the resolution of prediction value across a finer irradiance space, you can use `?highRes_piPred()` function. List of models are also listed in `?Model_piCurve()`.

## Phytoplankton Absorption Coefficient

Chlorophyll-a–specific phytoplankton absorption, $a^*_{ph}(\lambda)$, and the corresponding phytoplankton absorption coefficient, $a_{ph}(\lambda)$, can be estimated from chl-a concentration using the empirical spectral parameterization of Bricaud et al. (1995). This approach provides a globally averaged relationship between chlorophyll concentration and phytoplankton absorption in the absence of direct optical measurements (Details `?absCoef`). Below is an example illustrating the computation and visualization of $a_{ph}(\lambda)$ for a chl-a concentration of 0.9 mg m$^{-3}$. 
 
```{r}
absCoef(chla = 0.9) |>
    ggplot(aes(x = lambda_nm, y = a_ph)) +
    geom_line() +
    theme_bw(base_size = 14) +
    labs(
        x = "Wavelength (nm)",
        y = expression(a[ph](lambda)~~(m^{-1}))
    )
```


## Photosynthetically Available Radiation

Photosynthetically available radiation (PAR) is estimated by integrating downwelling spectral irradiance across the 400–700 nm photosynthetically active waveband. In **piCurve**, sea-surface spectral irradiance is computed using a Gregg–Carder–type parameterization with climatological atmospheric inputs (details: `?sea_surface_irradiance_piCurve`), providing a practical approach for estimating PAR in the absence of direct irradiance measurements. Below is an example illustrating the computation and visualization of PAR for a given date and solar zenith angle. 

```{r}
PAR_sea_surface_piCurve(
  zenith_angle = 30,
  database_info = GreggCarder1990,
  sample_date = "2024-07-24"
)
```


When not provided, the solar zenith angle can be computed from geographic coordinates (latitude and longitude) and time using `?solar_zenith_angle_piCurve`. Building on this, the figure below illustrates the global distribution of sea-surface PAR at local noon on March 31, 2026.

```{r}
Date <- as.Date("2026-03-31")

lat <- 0
long <- 0

sza <- 
    solar_zenith_angle_piCurve(date = Date, hour_utc = 12, latitude = lat, longitude = long)

PAR_sea_surface_piCurve(zenith_angle = sza, GreggCarder1990, sample_date = Date)
```

```{r}

#--------------------------------------------------
# 1. Create global grid
#--------------------------------------------------
df_global <- 
    expand.grid(
        longitude = seq(-180, 180, by = 2),
        latitude  = seq(-90, 90, by = 2)
        ) |>
    as_tibble()

#--------------------------------------------------
# 2. Approximate UTC time corresponding to local noon
#    local time = UTC + longitude/15
#    so local noon => UTC = 12 - longitude/15
#--------------------------------------------------
df_global <- 
    df_global |>
    # mutate(hour_utc = 12 - longitude / 15)
    mutate(hour_utc = (12 - longitude / 15) %% 24)

#--------------------------------------------------
# 3. Compute solar zenith angle
#--------------------------------------------------
df_global <-
    df_global |>
    mutate(sza = mapply(
        FUN = function(lat, lon, hr) {
            solar_zenith_angle_piCurve(
                date      = Date,
                hour_utc  = hr,
                latitude  = lat,
                longitude = lon
            )
        },
        lat = latitude,
        lon = longitude,
        hr  = hour_utc
    ))

#--------------------------------------------------
# 4. Compute PAR
#    If sun is below horizon / zenith >= 90, set PAR = 0
#--------------------------------------------------
df_global <-
    df_global |>
    rowwise() |>
    mutate(PAR = ifelse(
        is.na(sza) || sza >= 90,
        0,
        PAR_sea_surface_piCurve(
            zenith_angle  = sza,
            database_info = GreggCarder1990,
            sample_date   = Date
        )$PAR[1]
    )) |>
    ungroup()

#--------------------------------------------------
# 5. Plot global PAR map
#--------------------------------------------------
world_map <- map_data("world")

ggplot() +
    geom_raster(data = df_global,
                aes(x = longitude, y = latitude, fill = PAR)) +
    geom_polygon(
        data = world_map,
        aes(x = long, y = lat, group = group),
        fill = NA,
        color = "black",
        linewidth = 0.2
    ) +
    coord_fixed(xlim = c(-180, 180),
                ylim = c(-90, 90),
                expand = FALSE) +
    scale_fill_viridis_c(
        name = expression(PAR ~ (W ~ m ^ -2)),
        limits = c(0, 500),
        breaks = seq(0, 500, by = 100),
        guide = guide_colorbar(
            direction = "horizontal",
            barwidth  = unit(0.7, "npc"),
            barheight = unit(0.03, "npc"),
            title.position = "top",
            label.position = "bottom"
        )
    ) +
    labs(
        title = "Global sea-surface PAR at local noon",
        subtitle = "March 31, 2026",
        x = "",
        y = ""
    ) +
    theme_bw(base_size = 13) +
    theme(legend.position = "bottom", axis.title = element_blank())
```

## Citation

This package is developed based on the research outlined in the following paper. If you utilize the `piCurve` package in your work, we kindly request that you cite the referenced publication.

M. Amirian, M., Finkel, Z.V., Devred, E., Irwin, A.J. "*Parameterization of photoinhibition for phytoplankton*". Commun Earth Environ 6, 707 (2025). https://doi.org/10.1038/s43247-025-02686-3

M. Amirian, M. and Irwin, A.J. "*piCurve: an R package for modeling photosynthesis–irradiance curves*". arXiv preprint arXiv:2508.14321v1, (2025). https://arxiv.org/abs/2508.14321
