---
title: "piCurve Package"
output:
  md_document:
    variant: markdown_github
---

## Overview

The purpose of **piCurve** package to fit a wide range of photosynthesis-irradiance (PI) models to data. The optimal parameters for each model are estimated through non-linear optimization using two statistical approaches: minimizing the mean squared error (MSE) and maximum likelihood esitmation (MLE).

## Installation

Install from github using

```
remotes::install_github("Mohammad-Amirian/piCurve") # provide auth_token while package is private
```

## How to use piCurve

```{r load, message=FALSE}
library(piCurve)
library(ggplot2)
library(viridis)
```

### Light-saturation models

We can use a model to create some simulated data. This code creates a data table with two variables: irradiance, `I`, and photosynthetic rate, `PP`, for irradiances from 0 to 300 µmol photons m-2 s-1 using the hyperbolic tangent model of Jassby & Platt (1976).
A list of models can be found in the documentation for `Model_piCurve()`. 

```{r generate-data-1}
    set.seed(987)  # Update the seed value to any integer
    params <- c(Pmax = 20, alpha = 0.6, R = 0) # model parameters
    df <- tibble::tibble(I = seq(0, 300, length = 40)) # generate an irradiance profile
    
    df$PP <- # generate data using Jassby's hyperbolic tangent model
        Model_piCurve(params, model_name = "tanh", data = df, data_type = "ls") +
        5 * rnorm(length(df$I), 0, 0.2)  # add some noise to PP
```

Here is the top of the table:

```{r, include=TRUE}
head(round(df, digits = 2), 5)
```

To find the optimal parameters for the selected model with the `MSE` method, specify `STATapp = "MSE"` or leave that empty. The default approach is `MSE`. 

```{r}
fit_mse <-
    OptPar_piCurve(
        parameters = params,
        model_name = "tanh",
        data = df,
        data_type = "ls"
    )

fit_mse
```

For `MLE` method, one simply needs to specify `STATapp = "MLE"`:

```{r}
fit_mle <-
    OptPar_piCurve(
        parameters = c(params, StDev = 2),
        model_name = "tanh",
        data = df,
        # Hessian = TRUE, # default is FALSE
        STATapp = "MLE",
        data_type = "ls"
    )

fit_mle
```

The information matrix can be computed when the model is fit by setting the option `Hessian = TRUE`, or later using the function `InfoMat_piCurve`:

```{r}
fit_mle$hessian <-
    InfoMat_piCurve(
        parameters = fit_mle$par,
        model_name = fit_mle$model,
        data = df,
        data_type = "ls"
    )

fit_mle$hessian 
```

Compute predicted values of photosynthetic rate and visualizing the fit to the data:

```{r}
df$P_est <-
    Model_piCurve(
        fit_mle$par,
        model_name = fit_mse$model,
        data_type = "ls",
        data = df
    )

df |> 
    ggplot(aes(I, PP)) + 
    geom_point(size = 2) + 
    geom_line(aes(x = I, y = P_est), linewidth = 1, alpha = 0.7) +
    theme_bw() +
    labs(x = "Irradiance", y = "Photosynthetic Rate")
```

#### Confidence Interval

To calculate confidence intervals for the parameters, the `MLE` method must be used with `Hessian = TRUE`. The function `ConfInt_piModel()` provides the standard errors and confidence intervals:

```{r}
ConfInt_piCurve(Fitted_Model = fit_mle)
```

### Photoinhibition models

Here is another example using simulated data that exhibit photoinhibition.
A list of models can be found in the documentation for `Model_piCurve()`. 

```{r generate-data-2}
set.seed(987)  # Update the seed value to any integer
params <- c(Pmax = 20, alpha = 0.8, beta = 0.1, R = 0) # model parameters
df <- tibble::tibble(I = seq(0, 300, length = 40)) # generate an irradiance profile

df$PP <- # generate data using Jassby's hyperbolic tangent model
    Model_piCurve(params, model_name = "2tanh", data = df, data_type = "ph") +
    5 * rnorm(length(df$I), 0, 0.2)  # add some noise to PP
head(df, 3)
```

Here we fit the double-tanh (Ph10) model to the simulated data:

```{r}
fit_mle <-
    OptPar_piCurve(
        parameters = c(params, StDev = 2),
        model_name = "2tanh",
        data = df,
        STATapp = "MLE",
        data_type = "ph"
    )

fit_mle
```

and we visualize the fit:

```{r}
df$P_est <-
    Model_piCurve(
        fit_mle$par,
        model_name = fit_mle$model,
        data_type = "ph",
        data = df
    )

df |> 
    ggplot(aes(I, PP)) + 
    geom_point(size = 2) + 
    geom_line(aes(x = I, y = P_est), linewidth = 1, alpha = 0.7) +
    theme_bw() +
    labs(x = "Irradiance", y = "Photosynthetic Rate")
```

## Citation

This package is developed based on the research outlined in the following paper. If you utilize the `piCurve` package in your work, we kindly request that you cite the referenced publication.

Mohammad Amirian, Emmanuel Devred, Zoe V. Finkel, Andrew J. Irwin. “*An improved parameterization of photoinhibition in phytoplankton photosynthesis-irradiance curves*,” xxx 2024.
