---
title: "piCurve Package"
output: html_notebook
---

## Overview

The **piCurve** package integrates a diverse range of photosynthesis irradiance (PI) models, providing a user-friendly environment for users to explore the effects of different irradiance levels on photosynthetic activity using statistical methods. These models have been rigorously tested and validated against experimental data, ensuring their reliability and accuracy. The optimal parameters of the each formulation get estimated through non-linear optimization using two statistical approaches: MSE and MLE.

Individual and general model fit...

## How to use piCurve

```{r load}
rm(list = ls())
devtools::load_all()
library(piCurve)
library(ggplot2)
library(viridis)
library(fBasics)
```


### NO Pohotoinhibitio Case

To see how the package works, I will first generate some data using one of the PI formulations, the Jassby hyperbolic tangent model. Then I will add some noise to the generated data sample and use that as my given database.

```{r generate data 1, include=FALSE}
set.seed(987)  # Update the seed value to any integer
params <- c(Pmax = 20, alpha = 0.6, R = 0) # model parameters
df <- tibble::tibble(I = seq(0, 300, length = 40)) # generate an irradiance profile

df$PP <- # generate PP using Jassby's hyperbolic tangent model
    Model_piCurve(parameters = params, model_name = "Eq8Jassby", data = df) +
    5 * rnorm(length(df$I), 0, 0.2)  # add some noise to PP

head(round(df, digits = 2), 3)
```

where `I` and `PP` stand for irradiance and primary production profile, respectively.

#### Fit individually  

To find the optimal parameters for the embed formulations with `MSE` method, one can specify `STATapp = "MSE"` or leave that empty. The default approach is set on `MSE`. See the following example.

```{r}
OptPIparams(parameters = params, model_name = "eq2_blackman", data = df)
```

For `MLE` method, one simply needs to specify `STATapp = "MLE"`. See the following example

```{r}
OptPIparams(parameters = c(params, StDev=2), model_name = "eq2_blackman", data = df, STATapp = "MLE")
```

#### Fit all at Once

For a given data sample, let's find the optimal parameters for all models and examine which model captures the highest variation.

```{r useful functions, echo=FALSE}

# generalized function to fit equations on data samples all at once
apply_Eq <-
    function(model_index, STATapp, Hessian = FALSE, Control = list(), data = data) {
        if (model_index == 1) {
            # equations 1 is a linear regression and does not have Pmax param
            params <- params[names(params) != "Pmax"]
            }
        
        if (model_index %in% c(9, 10)) {
            # equations 6 and 7 have one extra param
            params <- add_shape_par(params)
        }
        
        OptPIparams(
            parameters = params,
            model_name = model_index,
            data = data,
            STATapp,
            Hessian = Hessian,
            Control = Control
        )
    }

```

```{r}
params <- c(Pmax = 20, alpha = 0.6, R = 0) 

fit_model <- # fit all equations
    lapply(1:10, function(idx)apply_Eq(model_index = idx, data = df)) 

outputs <-   # combine all the outputs of each fit
    lapply(1:10, 
           function(idx) reshape_piFits(fitted_model = fit_model[[idx]])
           )|> 
    dplyr::bind_rows() 

outputs |> apply(2, function(x)round(x, digits = 2)) |> data.frame()
```

```{r include=TRUE}
best_model <- dplyr::filter(outputs, R2adj %in% max(outputs$R2adj))

cat("formulation NO.", toupper(best_model$Model), "achieves", round(best_model$R2adj, 2) * 100, "% accuracy, capturing the highest amount of variation in the given data sample.")
```


#### Model Selection

To select the best model, one may consider using AIC/AICc or BIC -- depending on the question of interest. To this end, one may follow the same procedure.

```{r}
params <- c(Pmax = 20, alpha = 0.6, R = 0, StDev = 2) 
fit_model <- # fit all equations
    lapply(1:10, function(x)apply_Eq(model_index = x, STATapp = "mle", 
                                    Hessian = "TRUE", 
                                    Control = list(maxit = 2000),
                                    data = df))

outputs <-   # combine all the outputs of each fit
    lapply(1:10,
           function(idx){reshape_piFits(fit_model[[idx]], STATapp = "mle")}) |>
    dplyr::bind_rows() 

outputs |> apply(2, function(x)round(x, digits = 2)) |> data.frame()
```

```{r include=TRUE}
best_model <- dplyr::filter(outputs, AICc %in% min(outputs$AICc))

cat("formulation NO.", toupper(best_model$Model), " with the accuracy of", round(best_model$R2adj, 2) * 100, "has the lowest AICc/BIC value, meaning that the model is the best equation explaining the given data sample")
```

#### Confidence Interval

To calculate CI, `MLE` method must be used with `Hessian = TRUE`. Then, one simply needs to use `ConfInt_piModel()` function as shown below.

```{r}
cal_CI <-
    lapply(1:10,
           function(idx) {
               c(Model = idx, ConfInt_piModel(Fitted_Model = fit_model[[idx]], Clevel = 0.95))
           }) |>
    dplyr::bind_rows() 

cal_CI
```


#### visualization 

To visualize all the results in one plot, one can also use `ggplot` as done bellow.


```{r}
data_fit <- 
    lapply(1:10,
       function(idx)
           Model_piCurve(
               parameters = fit_model[[idx]]$par,
               model_name = idx,
               data = df
           )) 

eqs_info <- c(3, 8, 7)

fit <- 
    lapply(eqs_info, #1:10, 
           function(idx)data.frame(Phat=data_fit[[idx]], I = df$I, Model=idx)) |>
    dplyr::bind_rows() |> dplyr::mutate(Model = as.character(Model))
```

```{r}
fit <-     
    rbind(
        fit |> filter(Model == 3) |> mutate(Model = "Baly"),
        fit |> filter(Model == 7) |> mutate(Model = "Webb"),
        fit |> filter(Model == 8) |> mutate(Model = "Jassby")
    )
```


```{r}


p <- 
    ggplot(df, mapping = aes(x=I, y = PP)) + 
    geom_line(fit, mapping = aes(x = I, y = Phat, col = Model), linewidth = 1) +
    geom_point(size = 3, alpha = 0.7) + 
    theme_linedraw() +
    theme(legend.position = "top") +
        labs(x = "Irradiance [ Watts / m2 ]", 
         y = "       Primary Production
         [ mg C / (mg chl_a) h-1 ]") + 
    theme(axis.text = element_text(size = 14), axis.title=element_text(size = 12))

p +
    # scale_color_brewer(palette = "Set3")
    scale_color_viridis(discrete = TRUE, option = "D")
# ggsave("pi-fit-01.png", width = 10, height = 5)
```




### Photoinhibitio Case

Similar to the previous part, I will first generate some data using one of the PI formulations, the 5-parameters-extended Blackman model. Then I will add some noise to the generated data sample and use that as my given database.

```{r generate data 2, include=FALSE}
set.seed(987)  # Update the seed value to any integer
params <- c(Pmax = 22, alpha = 0.9, beta = 0.08, R = 0) # model parameters
df <- tibble::tibble(I = seq(0, 500, length = 40)) # generate an irradiance profile

df$PP <- # generate PP using 5-parameters-extended Blackman model
    Model_piCurve(parameters = params, model_name = "Eq11", data = df) +
    5 * rnorm(length(df$I), 0, 0.2)  # add some noise to PP

head(round(df, digits = 2), 3)
```

where `I` and `PP` stand for irradiance and primary production profile, respectively.

#### Fit individually  

To find the optimal parameters for the embed formulations with `MSE` method, one can specify `STATapp = "MSE"` or leave that empty. The default approach is set on `MSE`. See the following example.

```{r}
params <- Approx_Ibreak_par(params, df)
fit <- OptPIparams(parameters = params, model_name = "eq12", data = df)
fit
```

#### Fit all at Once

For a given data sample, let's find the optimal parameters for all models and examine which model captures the highest variation.

```{r useful functions update, echo=FALSE}
# generalized function to fit equations on data samples all at once
apply_Eq <-
    function(model_index, STATapp, Hessian = FALSE, Control = list(), data = data) {
        if (model_index == 1) {
            # Eq14: Steele 1962 doesn't have beta param
            params <- params[names(params) != "beta"]
            
            }
        
        if (model_index %in% c(13, 23, 24, 27, 31)) {
            # equations with an extra shape parameter
            params <- add_shape_par(params)
        }
        
        if (model_index %in% c(12, 18)){
            # equations with Ibreak parameter
            params <- Approx_Ibreak_par(parameters = params, data = data)
            }
        
        OptPIparams(
            parameters = params,
            model_name = model_index,
            data = data,
            STATapp,
            Hessian = Hessian,
            Control = Control
        )
    }
```

```{r}
params <- c(Pmax = 22, alpha = 0.9, beta = 0.08, R = 0) # model parameters

fit_model <- # fit all equations
    lapply(11:34, function(idx)apply_Eq(model_index = idx, data = df)) 

outputs <-   # combine all the outputs of each fit
    lapply(1:length(fit_model), 
           function(idx) reshape_piFits(fitted_model = fit_model[[idx]])
           )|> 
    dplyr::bind_rows() 

outputs |> apply(2, function(x)round(x, digits = 2)) |> data.frame() 
```

```{r include=TRUE}
best_model <- dplyr::filter(outputs, R2adj %in% max(outputs$R2adj))

cat("formulation NO.", toupper(best_model$Model), "achieves", round(best_model$R2adj, 2) * 100, "% accuracy, capturing the highest amount of variation in the given data sample.")
```


#### Model Selection

To select the best model, one may consider using AIC/AICc or BIC -- depending on the question of interest. To this end, one may follow the same procedure.

```{r}
params <- c(Pmax = 22, alpha = 0.9, beta = 0.08, R = 0, StDev = 1) # model parameters
fit_model <- # fit all equations
    lapply(11:34, function(x)apply_Eq(model_index = x, STATapp = "mle", 
                                    Hessian = "TRUE", 
                                    Control = list(maxit = 2000),
                                    data = df))

outputs <-   # combine all the outputs of each fit
    lapply(1:length(fit_model),
           function(idx){reshape_piFits(fit_model[[idx]], STATapp = "mle")}) |>
    dplyr::bind_rows() #|> dplyr::mutate(index = 1:length(fit_model))

outputs |> apply(2, function(x)round(x, digits = 2)) |> data.frame() 
```

```{r include=TRUE}
best_model <- dplyr::filter(outputs, AICc %in% min(outputs$AICc))

cat("formulation NO.", toupper(best_model$Model), " with the accuracy of", round(best_model$R2adj, 2) * 100, "has the lowest AICc/BIC value, meaning that the model is the best equation explaining the given data sample")
```


#### visualization 

To visualize all the results in one plot, one can also use `ggplot` as done bellow.


```{r}
good_fits <- outputs |> dplyr::filter(R2 > 0.85) 

data_fit <- 
    lapply(good_fits$Model,
       function(idx)
           Model_piCurve(
               parameters = good_fits |> dplyr::filter(Model == idx) |> as.vector() |> unlist(),
               model_name = idx,
               data = df
           )
       )
```


```{r}
fit <- 
    lapply(1:length(data_fit), 
           function(idx)data.frame(Phat=data_fit[[idx]], I = df$I, Model= good_fits$Model[idx])) |>
    dplyr::bind_rows() |> dplyr::mutate(Model = as.character(Model))
```


```{r}
p <- 
    ggplot(df, mapping = aes(x=I, y = PP)) + 
    geom_line(fit, mapping = aes(x = I, y = Phat, col = Model), linewidth = 1) +
    geom_point(size = 3, alpha = 0.7) + 
    theme_linedraw() +
    theme(legend.position = "top") +
        labs(x = "Irradiance [ Watts / m2 ]", 
         y = "       Primary Production
         [ mg C / (mg chl_a) h-1 ]") + 
    theme(axis.text = element_text(size = 14), axis.title=element_text(size = 12))

p +
    scale_color_brewer(palette = "Set1")
    # scale_color_viridis(discrete = TRUE, option = "D")
    
```

