---
title: "piCurve Package"
output:
  md_document:
    variant: markdown_github
---

## Overview

The **piCurve** package integrates a diverse range of photosynthesis irradiance (PI) models, providing a user-friendly environment for users to explore the effects of different irradiance levels on photosynthetic rate using statistical methods. The models have been rigorously tested and validated against experimental data, ensuring their reliability and accuracy. The optimal parameters of the each formulation get estimated through non-linear optimization using two statistical approaches: MSE and MLE.

## How to use piCurve

```{r load}
rm(list = ls())
devtools::load_all()
library(piCurve)
library(ggplot2)
library(viridis)
library(fBasics)
```

### Light-saturation models

The following data sample is given where `I` and `PP` stand for irradiance and photosynthetic rate, respectively.

```{r generate data 1, include=FALSE}
set.seed(987)  # Update the seed value to any integer
params <- c(Pmax = 20, alpha = 0.6, R = 0) # model parameters
df <- tibble::tibble(I = seq(0, 300, length = 40)) # generate an irradiance profile

df$PP <- # generate data using Jassby's hyperbolic tangent model
    Model_piCurve(params, model_name = "tanh", data = df, data_type = "ls") +
    5 * rnorm(length(df$I), 0, 0.2)  # add some noise to PP
```

```{r, include=TRUE}
head(round(df, digits = 2), 5)
```

To find the optimal parameters for the embed formulations with `MSE` method, one can specify `STATapp = "MSE"` or leave that empty. The default approach is set on `MSE`. See the following example.

```{r}
fit_mse <-
    OptPar_piCurve(
        parameters = params,
        model_name = "tanh",
        data = df,
        data_type = "ls"
    )

fit_mse
```

For `MLE` method, one simply needs to specify `STATapp = "MLE"`. See the following example

```{r}
fit_mle <-
    OptPar_piCurve(
        parameters = c(params, StDev = 2),
        model_name = "tanh",
        data = df,
        # Hessian = TRUE,
        STATapp = "MLE",
        data_type = "ls"
    )

fit_mle
```

In case if, the user forgot to set `Hessian = TRUE`, the information matrix can be calculated using `InfoMat_piCurve` as shown below.

```{r}
fit_mle$hessian <-
    InfoMat_piCurve(
        parameters = fit_mle$par,
        model_name = fit_mle$model,
        data = df,
        data_type = "ls"
    )

fit_mle$hessian 
```

visualizing the fit

```{r}
df$P_est <-
    Model_piCurve(
        fit_mle$par,
        model_name = fit_mse$model,
        data_type = "ls",
        data = df
    )

df |> 
    ggplot(aes(I, PP)) + 
    geom_point(size = 2) + 
    geom_line(aes(x = I, y = P_est), linewidth = 1, alpha = 0.7) +
    theme_bw() +
    labs(x = "Irradiance", y = "Photosynthetic Rate")
```

#### Confidence Interval

To calculate CI, `MLE` method must be used with `Hessian = TRUE`. Then, one simply needs to use `ConfInt_piModel()` function as shown below.

```{r}
ConfInt_piCurve(Fitted_Model = fit_mle)
```

### Photoinhibition models

Similar to above, one can alter the codes to photoinhibition models. A list of models can be found under details in `?Model_piCurve()`. Consider the following data sample, let's fit double-tanh (Ph10) model on it.

```{r generate data 2, include=FALSE}
set.seed(987)  # Update the seed value to any integer
params <- c(Pmax = 20, alpha = 0.8, beta = 0.1, R = 0) # model parameters
df <- tibble::tibble(I = seq(0, 300, length = 40)) # generate an irradiance profile

df$PP <- # generate data using Jassby's hyperbolic tangent model
    Model_piCurve(params, model_name = "2tanh", data = df, data_type = "ph") +
    5 * rnorm(length(df$I), 0, 0.2)  # add some noise to PP
```


```{r}
head(df, 3)
```

```{r}
fit_mle <-
    OptPar_piCurve(
        parameters = c(params, StDev = 2),
        model_name = "2tanh",
        data = df,
        STATapp = "MLE",
        data_type = "ph"
    )

fit_mle
```

visualizing the fit

```{r}
df$P_est <-
    Model_piCurve(
        fit_mle$par,
        model_name = fit_mle$model,
        data_type = "ph",
        data = df
    )

df |> 
    ggplot(aes(I, PP)) + 
    geom_point(size = 2) + 
    geom_line(aes(x = I, y = P_est), linewidth = 1, alpha = 0.7) +
    theme_bw() +
    labs(x = "Irradiance", y = "Photosynthetic Rate")
```

## Citation

This package is developed based on the research outlined in the following paper. If you utilize the `piCurve` package in your work, we kindly request that you cite the referenced publication.

Mohammad Amirian, Emmanuel Devred, Zoe V. Finkel, Andrew J. Irwin. “*An improved parameterization of photoinhibition in phytoplankton photosynthesis-irradiance curves*,” xxx 2024.
